---
title: "Barcelona's historical OpenStreetMap network"
subtitle: "Street hierarchy and traffic-related attributes for noise-complaints research"
date: "2026-10-08"
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: true
    theme: cosmo
execute:
  warning: false
  message: false
---

## Overview

This report describes Barcelona's OpenStreetMap (OSM) road and
path network as mapped on **1 January 2017**.

The dataset retains all OSM `highway` categories, including
motor-vehicle roads, pedestrian streets, footways and cycleways.

The purpose is to assess the availability and suitability of
historical street-network attributes for research on noise
complaints.

```{r}
#| label: setup
#| include: false

library(sf)
library(dplyr)
library(readr)
library(leaflet)
library(DT)
library(ggplot2)

roads <- st_read(
  "data/barcelona_osm_roads_170101.gpkg",
  quiet = TRUE
)

boundary <- st_read(
  "data/barcelona_boundary.gpkg",
  quiet = TRUE
)

quality <- read_csv(
  "outputs/quality_summary.csv",
  show_col_types = FALSE
)

highway_summary <- read_csv(
  "outputs/highway_summary.csv",
  show_col_types = FALSE
)

completeness <- read_csv(
  "outputs/attribute_completeness_overall.csv",
  show_col_types = FALSE
)

# Broad categories for visualisation only.
# Original highway tags remain unchanged.

roads <- roads |>
  mutate(
    road_group = case_when(
      highway == "pedestrian" ~ "Pedestrian street",
      highway %in% c(
        "footway", "path", "steps", "bridleway",
        "corridor", "crossing", "platform"
      ) ~ "Pedestrian/path",
      highway == "cycleway" ~ "Cycleway",
      highway == "living_street" ~ "Living street",
      highway %in% c(
        "residential", "service", "unclassified",
        "tertiary", "tertiary_link",
        "secondary", "secondary_link",
        "primary", "primary_link",
        "trunk", "trunk_link",
        "motorway", "motorway_link"
      ) ~ "Motor-vehicle road",
      TRUE ~ "Other"
    )
  )

roads_ll <- st_transform(roads, 4326)
boundary_ll <- st_transform(boundary, 4326)
```

## Dataset summary

```{r}
#| echo: false

data.frame(
  Metric = c(
    "OSM snapshot",
    "Number of features",
    "Total mapped feature length (km)"
  ),
  Value = c(
    "2017-01-01",
    format(nrow(roads), big.mark = ","),
    round(sum(roads$length_m) / 1000, 1)
  )
) |>
  DT::datatable(
    rownames = FALSE,
    options = list(dom = "t", paging = FALSE)
  )
```

## Network composition

The table presents all 27 original OSM `highway` categories,
showing the number of features, total mapped length and
percentage of the network.

```{r}
#| echo: false

highway_summary |>
  mutate(
    length_km = round(length_km, 2),
    pct_length = round(pct_length, 1)
  ) |>
  DT::datatable(
    rownames = FALSE,
    filter = "top",
    options = list(pageLength = 15)
  )
```

The figure visualises the mapped lengths reported in the
table, allowing comparison across all categories.

```{r}
#| echo: false

highway_summary |>
  mutate(highway = reorder(highway, length_km)) |>
  ggplot(aes(x = length_km, y = highway)) +
  geom_col(fill = "#3973A8") +
  labs(
    x = "Mapped length (km)",
    y = NULL,
    title = "Network length by OSM highway category"
  ) +
  theme_minimal(base_size = 12)
```

## Interactive network map

The map shows the complete historical network. For readability,
the original OSM `highway` categories are combined into six
broader groups. This aggregation is used only for visualisation;
the original classifications remain available in the dataset
and in the map popups.

```{r}
#| echo: false

pal <- colorFactor(
  palette = c(
    "Motor-vehicle road" = "#444444",
    "Pedestrian street" = "#E69F00",
    "Pedestrian/path" = "#56B4E9",
    "Cycleway" = "#009E73",
    "Living street" = "#CC79A7",
    "Other" = "#999999"
  ),
  domain = roads_ll$road_group
)

leaflet(roads_ll, options = leafletOptions(preferCanvas = TRUE)) |>
  addProviderTiles(providers$Esri.WorldGrayCanvas) |>
  addPolylines(
    color = ~pal(road_group),
    weight = 2,
    opacity = 0.75,
    popup = ~paste0(
      "<b>Highway:</b> ", highway,
      "<br><b>OSM ID:</b> ", osm_id,
      "<br><b>Length:</b> ", round(length_m), " m"
    ),
    group = "Network"
  ) |>
  addPolylines(
    data = boundary_ll,
    color = "black",
    weight = 2,
    opacity = 0.8,
    fill = FALSE,
    group = "Municipal boundary"
  ) |>
  addLegend(
    position = "bottomright",
    pal = pal,
    values = roads_ll$road_group,
    title = "Network category"
  ) |>
  addLayersControl(
    overlayGroups = c("Network", "Municipal boundary"),
    options = layersControlOptions(collapsed = TRUE)
  )
```

The map groups correspond to the following original OSM categories:

| Map group | Original OSM `highway` categories |
|---|---|
| Motor-vehicle road | `residential`, `service`, `unclassified`, `tertiary`, `secondary`, `primary`, `trunk`, `motorway`, and their `_link` variants |
| Pedestrian street | `pedestrian` |
| Pedestrian/path | `footway`, `path`, `steps`, `bridleway`, `corridor`, `crossing`, `platform` |
| Cycleway | `cycleway` |
| Living street | `living_street` |
| Other | `track`, `raceway`, `construction`, `road` |

These groups are defined for visualisation and do not constitute
an official OSM classification. The motor-vehicle road group
does not imply that motor-vehicle access is permitted on every
segment.

## Pedestrian infrastructure

The map below focuses on pedestrian streets, footways,
paths and living streets.

```{r}
#| echo: false

pedestrian <- roads_ll |>
  filter(
    road_group %in% c(
      "Pedestrian street",
      "Pedestrian/path",
      "Living street"
    )
  )

leaflet(pedestrian, options = leafletOptions(preferCanvas = TRUE)) |>
  addProviderTiles(providers$Esri.WorldGrayCanvas) |>
  addPolylines(
    color = ~pal(road_group),
    weight = 2,
    opacity = 0.8,
    popup = ~paste0(
      "<b>Highway:</b> ", highway,
      "<br><b>OSM ID:</b> ", osm_id
    )
  ) |>
  addPolylines(
    data = boundary_ll,
    color = "black",
    weight = 2,
    opacity = 0.8,
    fill = FALSE
  ) |>
  addLegend(
    position = "bottomright",
    pal = pal,
    values = pedestrian$road_group,
    title = "Pedestrian-related categories"
  )
```

## Attribute completeness

The following chart compares the proportion of features with
a recorded attribute against the proportion of mapped length
covered by those features.

```{r}
#| echo: false

completeness |>
  select(attribute, pct_features, pct_length) |>
  tidyr::pivot_longer(
    cols = c(pct_features, pct_length),
    names_to = "measure",
    values_to = "percentage"
  ) |>
  mutate(
    measure = recode(
      measure,
      pct_features = "By feature count",
      pct_length = "By mapped length"
    )
  ) |>
  ggplot(
    aes(
      x = percentage,
      y = reorder(attribute, percentage),
      fill = measure
    )
  ) +
  geom_col(position = "dodge") +
  labs(
    x = "Attributes recorded (%)",
    y = NULL,
    fill = NULL,
    title = "Completeness of historical OSM attributes"
  ) +
  theme_minimal(base_size = 12)
```

## Methodological notes

### Historical representation

The dataset uses the OSM state as of 1 January 2017.
It does not necessarily represent every road or path
that physically existed at that date.

### Road hierarchy

The `highway` tag describes road function and classification,
not measured traffic volume.

### Missing attributes

Missing lane counts, speed limits, widths and other attributes
should not be interpreted as zero or as evidence of absence.

### Pedestrianisation

The `highway=pedestrian` category is not a validated inventory
of pedestrianised streets. Pedestrian-priority streets may also
appear under other tags.

### Network length

Lengths are sums of mapped OSM geometries. Overlapping
features may be counted more than once.

### Administrative boundary

The clipping boundary was obtained from current OSM data,
rather than reconstructed from a historical administrative
boundary.

### Data attribution

© OpenStreetMap contributors. OSM data is made available
under the Open Database License (ODbL).
